---
title: "Data Visualization"
author: "Matt Steele"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
---


```{r}
#| label: working directory
#| include: false
#| eval: false

setwd("<path_path_to_file_folder>")


```

<br />

```{r}
#| label: load tidyverse
#| eval: false
#| include: false

install.packages("tidyverse")
library(tidyverse)


```

<br />

------------------------------------------------------------------------

# Part 1: Introduction to GGPLOT2

visualize your data with ggplot

-   [ggplot package]("https://ggplot2.tidyverse.org/")

<br />

## Grammar of Graphics

ggplot allows you to visualize data by using three components: *a data set*, *a coordinate systems*, and *geoms*.

1.    Set the plot
2.    Set the coordinates
3.    Set the shape

<br />

------------------------------------------------------------------------

# Part 2: Setting the Plot and Coordinates

## Parameters

in ggplot you set you parameters or the variables you want to plot with
the **aesthetics** argument.

```{r}
#| label: aes
#| eval: false

# load scotus approval data

scotus <- read_csv("scotus_approval.csv")

# let's just view the pollster YouGov using the filter function

scotus_yg <- scotus |> 
  filter(pollster == "YouGov")

scotus_yg


# set the parameters and coordinates

scotus.line <- ggplot(scotus_yg, aes(date, per_yes))

# calling that object will givve us an empty plot


scotus.line


```

<br />

------------------------------------------------------------------------

# Part 3: Setting the plot shape

the **geom\_** or **stat\_** functions allows you to choose the shape.
We are going start by plotting a line graph so I will use
**geom_line()**.

[Plots Cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)


```         
We combine elements in ggplot using the (+) operator 
```

```{r}
#| label: geom_line
#| eval: false

scotus.line + 
  geom_line()
  

```

<br />

------------------------------------------------------------------------

# Part 4: Adding additional features

<br />


## Change color, size, and transparency of elements

|                           |                     |
|---------------------------|---------------------|
| **fill =** or **color =** | change colors       |
| **size =**                | change size         |
| **alpha =**               | change transparency |

```{r}
#| label: color, shape, transparency
#| eval: false

# add color

scotus.line + 
  geom_line(color = "coral")

# change the size

scotus.line + 
  geom_line(color = "coral", size = 2)


```

<br />

---


## Labeling your plot

the [labs element](https://ggplot2.tidyverse.org/reference/labs.html?q=labs) will allow you to add or change labels in the plot

```{r}
#| label: labs
#| eval: false

scotus.line + 
  geom_line(color = "coral", size = 2) +
  labs(
    title = "SCOTUS Approval",
    subtitle = "2023",
    caption = "polls from YouGov",
    y = "Approval",
    x = NULL
  )


```

<br />

---

## Built-in Themes

ggplot has [built-in themes](https://ggplot2.tidyverse.org/reference/index.html#themes) with pre-set settings for you 


```{r}
#| label: built-in theme
#| eval: false

scotus.line + 
  geom_line(color = "coral", size = 2) +
  labs(
    title = "SCOTUS Approval",
    subtitle = "2023",
    caption = "polls from YouGov",
    y = "Approval",
    x = NULL
  ) +
  theme_minimal()

```


<br />

---


## Customizing themes

the [theme element](https://ggplot2.tidyverse.org/reference/theme.html) will allow you to customize the appearance of axes, legends, and labels

```{r}
#| label: customize theme
#| eval: false

scotus.line + 
  geom_line(color = "coral", size = 2) +
  labs(
    title = "SCOTUS Approval",
    subtitle = "2023",
    caption = "polls from YouGov",
    y = "Approval",
    x = NULL
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, color = "navy"))

```


<br />

---

## Scales

the [scales element](https://ggplot2.tidyverse.org/reference/index.html#scales) allows you to fine-tune and adjust the mapping/scale of labels, breaks, and legends

- the **scale_x_date** element allows you to adjust your date elements on the x axis

- [Date Formats - strptime](https://rdrr.io/r/base/strptime.html)

```{r}
#| label: scales
#| eval: false

scotus.line + 
  geom_line(color = "coral", size = 2) +
  labs(
    title = "SCOTUS Approval",
    subtitle = "2023",
    caption = "polls from YouGov",
    y = "Approval",
    x = NULL
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, color = "navy")) +
  scale_x_date( date_breaks = "6 weeks",
                date_labels = "%b %d")
  
```

<br />

---

## Smoothed Lines

You can reduce overplotting using **loess** or **linear regression lines** with the [geom_smooth or stat_smooth element](https://ggplot2.tidyverse.org/reference/geom_smooth.html)

```{r}
#| label: geom_smooth
#| eval: false

scotus.line + 
  geom_smooth(color = "coral", size = 2) +
  labs(
    title = "SCOTUS Approval",
    subtitle = "2023",
    caption = "polls from YouGov",
    y = "Approval",
    x = NULL
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 20, color = "coral")) +
  scale_x_date( date_breaks = "6 weeks",
                date_labels = "%b %d")

```



<br />

---



## Export your plot

The [ggsave function](https://ggplot2.tidyverse.org/reference/ggsave.html) will export the most recent plot called in a fille type specified by the user


```{r}
#| label: export
#| eval: false

ggsave("scotus_approval.png", plot = my_plot, width = 6, height = 4, dpi = 300)


```

Additionally you can use the export options in RStudio's Plot tab in the Misc Pane

<br />

---


# Histogram

the [histogram plot](https://ggplot2.tidyverse.org/reference/geom_histogram.html) allows you to see the distribution of a continuous (dbl or num) variable

```{r}
#| label: histogram
#| eval: false


# load demographics data frame

demo <- read_csv("demographics.csv")


# let's look at the distribution of the age variable by creating a histogram

demo.hist <- ggplot(demo, aes(age))

demo.hist +
  geom_histogram()


```

<br />

---

## Binning

the binning argument allows you to group continuous data into discrete intervals or bins

```{r}
#| label: bin
#| eval: false

# number of bins to use

demo.hist +
  geom_histogram(bins = 10)

# length of a bins

demo.hist +
  geom_histogram(binwidth = 15)

```

<br />

---

## Change color, size, and transparency of elements

|                           |                     |
|---------------------------|---------------------|
| **fill =** or **color =** | change colors       |
| **size =**                | change size         |
| **alpha =**               | change transparency |



```{r}
#| label: theme options
#| eval: false



demo.hist +
  geom_histogram(bins = 25, color = "coral", fill = "skyblue", alpha = .5) +
  theme_light()

```

<br />

---

## Order of Elements

The order that the elements appear on the plot is dictated by its position in your code. 

- The first elements in the code appear at the bottom of the plot and the last elements appear on the top of you plot

<br />

---


## Reference line 

the [geom_vline/geom_hline element](https://ggplot2.tidyverse.org/reference/geom_abline.html) allows you to add a reference line to your plot


```{r}
#| label: vline
#| eval: false

# add a reference line

demo.hist +
  geom_vline(xintercept = 40, color = "navy", size = 3) +
  geom_histogram(bins = 25, color = "coral", fill = "skyblue", alpha = .5) +
  theme_light()

```


<br />

---

## Faceting

the [facet_grid](https://ggplot2.tidyverse.org/reference/facet_grid.html) or [facet_wrap](https://ggplot2.tidyverse.org/reference/facet_wrap.html) element will allow you to break your plot out by categorical variables

```{r}
#| label: facets
#| eval: false

demo.hist +
  geom_vline(xintercept = 40, color = "navy", size = 3) +
  geom_histogram(bins = 25, color = "coral", fill = "skyblue", alpha = .5) +
  theme_light() + 
  facet_wrap(facets = vars(inccat), nrow = 3)

```

<br />

------------------------------------------------------------------------

# Bar Graph

the [geom_bar element](https://ggplot2.tidyverse.org/reference/geom_bar.html?q=geom_bar) allows you create a bar chart uses the number of cases of each group in a categorical variable

```{r}
#| label: geom_bar
#| eval: false


demo.bar <- ggplot(demo, aes(carcat))

demo.bar +
  geom_bar()


```


<br />


the [geom_col element](https://ggplot2.tidyverse.org/reference/geom_bar.html?q=geom_bar) allows you to create a bar chart using a categorical and continuous variable


```{r}
#| label: geom_col
#| eval: false

demo.col <- ggplot(demo, aes(carcat, income))

demo.col +
  geom_col()

```

<br />

---

## Reorder Plot

you can order the bar graph using the [fct_reorder function](https://forcats.tidyverse.org/reference/fct_reorder.html) from Forcats

```{r}
#| label: fct_reorder
#| eval: false


demo.col <- ggplot(demo, aes(fct_reorder(carcat, income), income))

demo.col +
  geom_col()

```

<br />

---

## Add Additional Variable

you can use the fill argument in aes to map an additional variable onto individual bars

```{r}
#| label: aes fill
#| eval: false

demo.col +
  geom_col(aes(fill = ed))
  

```

<br />

---

## Add Color Palletes

The scale_fill_brewer function will allow you to add pre-built palettes to your plot

-   [Color Brewer](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)


```{r}
#| label: color brewer
#| eval: false

demo.col +
  geom_col(aes(fill = ed)) +
  scale_fill_brewer(palette = "Pastel1")

  

```

<br />

---


## Scales

the [scales element](https://ggplot2.tidyverse.org/reference/index.html#scales) allows you to fine-tune and adjust the mapping/scale of labels, breaks, and legends

- the **scale_y_continuous** or **scale_x_continuous** along with [label_number](https://scales.r-lib.org/reference/label_number.html) elements allows you to adjust a numeric axis

```{r}
#| label: scale continous
#| eval: false

demo.col +
  geom_col(aes(fill = ed)) +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(labels = scales::label_number_si())
  

```



